Molecular dynamics simulation of synchronization in driven particles

### 2020-11-17 manuscript/manuscript_ajp_2020.tex
added detail to single particle system

The simulation is controlled by a $for()$ loop
which runs from an intial to maximum $time$ integer.
Each time step represents a simulation time of $\Delta t=0.001$.
At each timestep
we evaluate the net force on each particle as a function of its position
$\vec{r}_i(t)$
and then use the equation of motion to move particles
to an updated position
$\vec{r}_i(t+\Delta t)$.
%
We model the 
particle dynamics 
with an overdamped
equation of motion
integrated with the Verlet method.
Here 
the suspending fluid is highly viscous
and exerts a damping force on the particles equivalent
to a linear drag force.
%Overdamped dynamics is a common assumption in models of colloidal particles.
%This is an assumption that 
The damping comes from a viscous fluid model
providing a nonconservative force,
modeled as a linear friction $\vec{F}_{drag} = -\eta \vec{v}_i$
sufficient so that
the acceleration of the particle is zero.
Such a model is appropriate when the particles
are small and the viscosity is high [finish... reference...]
%[This model is encountered in Ch. 2 of Taylor's Classical Mechanics....]

A single particle has the equation of motion 
\begin{equation}
\eta \vec{v}_i = \vec{F}_{l,i} + \sum_{i \neq j}^{N} \vec{F}_{ij} + \vec{F}_{D}(t).
\end{equation}
where $\eta = 1$.
The equation of motion provides a direct calculation of the velocity
of an individual particle from its location in the simulation
with respect to other particles $j$.
Since we model the acceleration as zero,
the Verlet method simplifies to 
the Euler method,
which is used to calculate the position at subsequent time steps.
%Most undergraduate students learn the Euler method in Computational Physics and Differential Equations, but don't encounter the Verlet method.
\begin{equation}
\vec{r}_i(t+\Delta t) = \vec{v}_i(t) \Delta t + \vec{r}_i(t).
\end{equation}

@@

We place a single particle in a periodic landscape with
corrugations along the y-direction 
\begin{equation}
  \label{eq:landscape}
  V_l(y) = V_{0y} \cos{(N_p \pi y / L)} 
\end{equation}
where $N_p=20$ are the number of troughs in the landscape,
and $V_{0y}=0.2$.
This is illustrated in Fig.~\ref{fig:1}(a) where
the red (blue) regions show local maxima (minima).
The code for generating
a two dimensional colored plot
of the potential landscape
is calculated by evaluating
the analytic function in Eq.~\ref{eq:landscape}
for a grid of values $(x_n,y_n)$.
%is included in the supplementary material,

%single DC case with figure

%increased DC
We apply an external applied force, 
with 
a constant $F_{AC}$ with frequency $\omega = 2\pi f$
then slowly increase $F_{DC}$
at a rate of [].
The particle hops between troughs in the energy landscape
when the superposition of $F_{AC}$ and $F_{DC}$ is large enough
to overcome the barrier height.
By modifying $F_{DC}$ we 
achieve a variety of oscillation modes.
A mode is a periodic pattern of hops
with a constant average particle velocity, $\bar{v}_{y}$
over a range of driving forces $F_{DC}$.
We illustrate mode-locking in 
the velocity-force plot in Fig.~\ref{fig:1}(b).
Here $\bar{v}_{y}$ is increasing in non-uniform steps,
with a quantized height of
$\bar{v}_{y} = n \lambda f$,
where $n$ is an integer,
$\lambda = S_Y/N_p = 36.5/20 = 1.825$ is the spatial period, or wavelength
of the landscape,
and $f = 0.01$ cycles per time unit.

@@

We illustrate the hopping pattern in Fig.~\ref{fig:1}(c)
and 
show the dynamics 
in supplementary materials \cite{supp1}.

>>>>

The simulation is controlled by a $for()$ loop
which runs from an initial to maximum $time$ integer.
Each integer timestep
represents a simulation time of $\Delta t=0.001$.
At each timestep
we evaluate the net force on each particle as a function of its position
$\vec{r}_i(t)$
and then integrate
the equation of motion to move particles
to an updated position
$\vec{r}_i(t+\Delta t)$.
%
The integration is simple
because 
we model the 
particle dynamics 
with an overdamped
equation of motion.
%integrated with the Verlet method.
Here 
the suspending fluid is highly viscous
and exerts a damping force on the particles equivalent
to a linear drag force.
%Overdamped dynamics is a common assumption in models of colloidal particles.
%This is an assumption that 
The damping comes from a viscous fluid model
providing a nonconservative force,
modeled as a linear friction $\vec{F}_{drag} = -\eta \vec{v}_i$
sufficient so that
the acceleration of the particle is zero.
Such a model is appropriate when the particles
are small and the viscosity is high [finish... reference...]
This model is encountered in standard texts,
for instance Ch. 2 of Taylor's Classical Mechanics.

A particle has the equation of motion 
\begin{equation}
\eta \vec{v}_i = \vec{F}_{l,i} + \sum_{i \neq j}^{N} \vec{F}_{ij} + \vec{F}_{D}(t).
\end{equation}
where $\eta = 1$.
The equation of motion provides a direct calculation of the velocity
of an individual particle from its location in the simulation
with respect to other particles $j$.
Since we model the acceleration as zero,
the Verlet method simplifies to 
the Euler method,
which is used to calculate the position at subsequent time steps.
%Most undergraduate students learn the Euler method in Computational Physics and Differential Equations, but don't encounter the Verlet method.
\begin{equation}
\vec{r}_i(t+\Delta t) = \vec{v}_i(t) \Delta t + \vec{r}_i(t).
\end{equation}

@@

We drive a single particle across a periodic landscape 
along the y-direction 
\begin{equation}
  \label{eq:landscape}
  V_l(y) = V_{0y} \cos{(N_p \pi y / L)} 
\end{equation}
with $N_p=20$ troughs in the landscape.
%and $V_{0y}=0.2$.
This is illustrated in Fig.~\ref{fig:1}(a) where
the red (blue) regions show local maxima (minima).
The code for generating
a two dimensional colored plot
of the potential landscape
is calculated by evaluating
the analytic function in Eq.~\ref{eq:landscape}
for a grid of values $(x_n,y_n)$.
%is included in the supplementary material,
The numerical implementation of the landscape 
is calculated with
\begin{equation}
  \label{eq:landscape}
  F_y(y) = F_{0y} \sin{(N_p \pi y / L)} 
\end{equation}
where $F_{0y} = 0.1$.
We drive the particle with
$F_{AC}=0.2$, $F_{DC}=0.01$, and $f=0.01$ cycles per time unit.
When the superposition of $F_{AC}$ and $F_{DC}$ is large enough
to overcome the barrier height of the landscape troughs,
the particle hops between troughs in the energy landscape.
In Fig.~\ref{fig:0} we
show the relationship between the applied force and particle
position.
In Fig.~\ref{fig:0}(a)
we plot the applied driving force
as a function of time
and in Fig.~\ref{fig:0}(b) 
we show the position of the particle
as a function of time.
The initial position is near $y=0$
and the initial driving force is minimum
$F_D(t=0) = F_{DC}$.
Over the time period $T/2 = 50$ %time units
the particle moves
in the positive y-direction
while $F_{D}(t) > 0 $
through three
substrate troughs, reaching a maximum of $y/\lambda = 3$.
When $F_{D}(t) < 0 $
the particle moves in the negative y-direction
reaching a minimum position
of $y/\lambda = 1$
over the period $T = 1/f = 100$.
The average velocity $\bar{v}_y$
is the average displacement $\Delta y = y(t_0+T) - y(t_0)$
over the period of the driving force.
In Fig. ~\ref{fig:0}
the displacement is a single wavelength
of the substrate $y(t_0+T) - y(t_0) = \lambda$.
Thus the average velocity 
is $\bar{v}_y = \lambda f$, where $T = 1/f$.

%single DC case with figure
%dc_current 0.01
%ac_current 0.2
%ac_frequency 0.0001
%fp_y 0.1
\begin{center}
\begin{figure}[h!]
\centering
\includegraphics[scale=.25]{position}
\caption{The position as a function of time of a single driven particle
  normalized by the period of the substrate $\lambda$.
  }
\label{fig:0}
\end{figure}
\end{center}

The hopping pattern of the driven particle is periodic,
and could be achieved with a variety of
driving forces.  %, as we will demonstrate in Fig.~\ref{fig:1}
In Fig.~\ref{fig:1}
we show the result increasing the 
force to the particle as a function of time.  
%increased DC
We apply an external applied force, 
with 
a constant $F_{AC}$ with frequency $\omega = 2\pi f$
then slowly increase $F_{DC}$
at a rate of $\Delta F_{DC} = 0.001$ every $\Delta t = 4000$ integer timesteps
or $40$ time units.
%this seems fast, but it is slow enough to estimate
%the velocity accurately.
%this is not be so for a multiple particle system!
By modifying $F_{DC}$ we 
achieve a variety of oscillation modes.
A mode is a periodic pattern of hops
with a constant average particle velocity, $\bar{v}_{y}$
over a range of driving forces $F_{DC}$.
We illustrate mode-locking in 
the velocity-force plot in Fig.~\ref{fig:1}(b).
Here $\bar{v}_{y}$ is increasing in non-uniform steps,
with a quantized height of
$\bar{v}_{y} = n \lambda f$,
where $n$ is an integer,
$\lambda = S_Y/N_p = 36.5/20 = 1.825$ is the spatial period, or wavelength
of the landscape,
and $f = 0.01$ cycles per time unit.

@@

We illustrate the hopping pattern in Fig.~\ref{fig:1}(b)
and 
show the dynamics 
in supplementary materials \cite{supp1}.
